Something about cetacean heart rates and ballistocardiograms

### 2021-09-03 analysis/paper/paper.Rmd
Refactor paper to incorporate changes in filtering steps

```{r process-acc}
acc_400hz <- corky_400hz
fs <- 400
fny <- fs / 2
acc_filter <- signal::butter(4, c(1.0, 25.0) / fny, type = "pass") 
apply_filter <- function(x) signal::filtfilt(acc_filter, x)
acc_400hz <- acc_400hz %>% 
  mutate(across(c(surge, sway, heave), apply_filter, .names = "{.col}_filt"))

# This window (in seconds) will be used for smoothing multiple signals and in
# heartbeat detection
window <- 0.25
prh_10hz <- corky_10hz %>% 
  mutate(jerk_smooth = tma(jerk, window * 10))
```

```{r align-ecg}
acc_400hz$ecg_beat <- FALSE
beat_idx <- approx(acc_400hz$dt, 
                   seq_along(acc_400hz$dt), 
                   xout = ecg_beats$dt,
                   method = "constant")$y %>% 
  na.omit()
acc_400hz$ecg_beat[beat_idx] = TRUE

prh_10hz$ecg_beat <- FALSE
beat_idx <- approx(prh_10hz$dt, 
                   seq_along(prh_10hz$dt), 
                   xout = ecg_beats$dt,
                   method = "constant")$y %>% 
  na.omit()
prh_10hz$ecg_beat[beat_idx] = TRUE
```

```{r bcg-beats}
acc_400hz$surge_range <- local_range(acc_400hz$surge_filt, window * fs)
acc_400hz$bcg_beat <- find_beats_lsr(acc_400hz$surge_filt, window, fs)
acc_400hz$ecg_bpm[acc_400hz$ecg_beat] <- with(
  filter(acc_400hz, ecg_beat), 
  1 / as.numeric(lead(dt) - dt, unit = "mins")
)
acc_400hz$bcg_bpm[acc_400hz$bcg_beat] <- with(
  filter(acc_400hz, bcg_beat), 
  1 / as.numeric(lead(dt) - dt, unit = "mins")
)

prh_10hz$bcg_beat <- FALSE
bcg_idx <- approx(prh_10hz$dt, 
                  seq_along(prh_10hz$bcg_beat), 
                  acc_400hz$dt[acc_400hz$bcg_beat],
                  "constant")$y
prh_10hz$bcg_beat[bcg_idx] <- TRUE
```

```{r bcg-figs}
ca_tz <- "Etc/GMT+7"
bcg_begin <- lubridate::ymd_hms("2021-08-16 10:09:24.5", tz = ca_tz)
bcg_end <- lubridate::ymd_hms("2021-08-16 10:09:39", tz = ca_tz)
bcg_sample <- filter(acc_400hz, between(dt, bcg_begin, bcg_end))
bcg_sample_10hz <- filter(prh_10hz, between(dt, bcg_begin, bcg_end))

@@

```{r corky-bcg-ecg, fig.cap="A: BCG, B: ECG, C: BPM comparison"}
# A: Surge + LSR
bcg_surge_lsr <- bcg_sample %>% 
  rename(Surge = surge_filt, LSR = surge_range) %>% 
  pivot_longer(c(Surge, LSR), 
               names_to = "var",
               values_to = "val")

@@

# C: BCG + ECG
pC <- ggplot(filter(bcg_sample, bcg_beat), aes(dt, bcg_bpm)) +
  geom_line(size = 0.2) +
  geom_line(aes(y = ecg_bpm), filter(bcg_sample, ecg_beat)) +
  geom_point(color = "blue") +
  geom_point(aes(y = ecg_bpm), filter(bcg_sample, ecg_beat), color = "red") +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = "Heart rate (bpm)") +
  expand_limits(x = range(bcg_sample$dt) + c(0, 1), y = c(30, 70)) +
  t

@@

```{r corky-bcg-acc, fig.cap="A: The BCG signal is clearest in surge (longitudinal axis acceleration). Solid line: surge bandpass filtered to [1.0 25.0] Hz. Dashed line: local range of surge in 0.25 s sliding window (smoothed). Blue dots indicate peak of the BCG I-wave. B: BCG signal is not as strong in heave (dorso-ventral axis) or sway (lateral axis) as in surge. The sway signal is likely due to movement of the chest cavity (see video S1). C: Some, but not all, heart beats are visible as peaks in the norm of the jerk vector."}
# A: Surge + LSR
# pA as above

# B: Sway, heave
bcg_sway_heave <- bcg_sample %>% 
  rename(Sway = sway_filt, Heave = heave_filt) %>% 
  pivot_longer(c(Sway, Heave), 
               names_to = "axis",
               values_to = "acceleration")
pB <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
  geom_line(size = 0.2) +
  geom_point(data = filter(bcg_sway_heave, bcg_beat), color = "blue") +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  facet_grid(axis ~ .) +
  labs(y = bquote(Acc~(m~s^{-2}))) +
  t

# C: Jerk
pC <- ggplot(bcg_sample_10hz, aes(dt, jerk_smooth)) +
  geom_line(size = 0.2) +
  geom_point(data = filter(bcg_sample_10hz, bcg_beat), color = "blue") +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = bquote(Jerk~(m~s^{-3}))) +
  expand_limits(y = 0) +
  t

cowplot::plot_grid(pA, pB, pC,
                   align = "v", 
                   axis = "lr",
                   nrow = 3, 
                   # rel_heights = c(1, 1, ),
                   labels = "AUTO")
```

```{r bw-data}
fs <- 400
fny <- fs / 2
acc_filter <- signal::butter(4, c(1.5, 7.5) / fny, type = "pass") 
apply_filter <- function(x) signal::filtfilt(acc_filter, x)
window <- 1.8 #s

bw_elg <- bw180905_53_elg

bw_400hz <- bw180905_53_400hz %>% 
  mutate(
    across(c(surge, sway, heave), apply_filter, .names = "{.col}_filt"),
    rid_left = approx(bw_elg$start, bw_elg$region_id, dt, "constant")$y,
    rid_right = approx(bw_elg$stop, bw_elg$region_id, dt, "constant", yleft = 0)$y + 1,
    region_id = ifelse(rid_left == rid_right, rid_left, NA)
  ) %>% 
  select(-c(rid_left, rid_right))

bw_10hz <- bw180905_53_10hz %>% 
  mutate(jerk = njerk(aw, 10),
         jerk_smooth = tma(jerk, window * 10))
```

```{r bw-beats}
bw_400hz_beats <- bw_400hz %>% 
  mutate(jerk = approx(bw_10hz$dt, bw_10hz$jerk_smooth, bw_400hz$dt, rule = 2)$y) %>% 
  filter(!is.na(region_id)) %>%
  group_by(region_id) %>% 
  mutate(bcg_beat = find_beats_jerk(jerk, surge_filt, window, fs),
         bcg_bpm = bpm(bcg_beat, dt)) %>% 
  ungroup() %>% 
  select(dt, bcg_beat, bcg_bpm)

bw_400hz <- bw_400hz %>% 
  mutate(surge_range = local_range(surge_filt, window * fs)) %>% 
  left_join(bw_400hz_beats, by = "dt")

bw_10hz$bcg_beat <- FALSE
bcg_idx <- approx(bw_10hz$dt, 
                  seq_along(bw_10hz$bcg_beat), 
                  bw_400hz$dt[bw_400hz$bcg_beat],
                  "constant")$y
bw_10hz$bcg_beat[bcg_idx] <- TRUE
```

```{r bw-bcg}
bw_bcg_plot <- function(rid) {
  ca_tz <- "Etc/GMT+7"
  bcg_sample <- filter(bw_400hz, region_id == rid)
  bcg_sample_10hz <- bw_10hz %>% 
    filter(between(dt, bcg_sample$dt[1], bcg_sample$dt[nrow(bcg_sample)]))

@@

  p1 <- ggplot(bcg_surge_lsr, aes(dt, val, linetype = var)) +
    geom_line(size = 0.2) +
    geom_point(data = filter(bcg_surge_lsr, bcg_beat), color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    scale_linetype_manual(values = c(Surge = "solid", LSR = "dashed")) +
    labs(y = bquote(Surge~(m~s^{-2}))) +
    t +
    theme(legend.position = "none")

@@

  p2 <- ggplot(filter(bcg_sample, bcg_beat), aes(dt, bcg_bpm)) +
    geom_line() +
    geom_point(color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    labs(y = "Heart rate (bpm)") +
    expand_limits(x = range(bcg_sample$dt), y = 0) +
    t

@@

  p3 <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
    geom_line(size = 0.2) +
    geom_point(data = filter(bcg_sway_heave, bcg_beat), color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    facet_grid(axis ~ .) +
    labs(y = bquote(Acc~(m~s^{-2}))) +
    t

@@

  p4 <- ggplot(bcg_sample_10hz, aes(dt, jerk_smooth)) +
    geom_line(size = 0.2) +
    geom_point(data = filter(bcg_sample_10hz, bcg_beat), color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    labs(y = bquote(Jerk~(m~s^{-3}))) +
    expand_limits(y = 0) +
    t

  p <- cowplot::plot_grid(p1, p2, p3, p4, 
                          align = "v", 
                          axis = "lr",
                          nrow = 4, 
                          rel_heights = c(1.5, 0.75, 1, 1),
                          labels = "AUTO")

@@

Figure \@ref(fig:corky-bcg-ecg) shows how we can have a caption and cross-reference for a plot.

>>>>

```{r process-acc}
fs <- 400
fny <- fs / 2
window <- 0.25 #s
bandpass_filter <- function(x, pass) {
  signal::butter(4, pass / fny, type = "pass") %>% 
    signal::filtfilt(x)
}
jerk_filter <- function(x, size) {
  signal::sgolayfilt(x, p = 4, n = size + (size %% 2) + 1, m = 1)
}
oo_400hz <- corky_400hz %>% 
  mutate(
    # Pass-band filter acceleration
    across(surge:heave, 
           bandpass_filter, pass = c(1.0, 25.0), 
           .names = "{.col}_filt"),
    # Detect heart beats using BCG's local surge range
    surge_range = local_range(surge_filt, window * fs),
    bcg_beat = find_beats_lsr(surge_filt, window, fs),
    bcg_bpm = bpm(bcg_beat, dt),
    # Calculate jerk norm and apply smoother
    njerk = njerk(cbind(surge_filt, sway_filt, heave_filt), 
                  p = 4, n = window * fs),
    njerk_smooth = tma(njerk, window * fs)
  )
```

```{r align-ecg}
# Align ECG-detected heart beats
oo_400hz$ecg_beat <- FALSE
beat_idx <- approx(oo_400hz$dt, 
                   seq_along(oo_400hz$dt), 
                   xout = ecg_beats$dt,
                   method = "constant")$y %>% 
  na.omit()
oo_400hz$ecg_beat[beat_idx] = TRUE
oo_400hz$ecg_bpm <- with(oo_400hz, bpm(ecg_beat, dt))
```

```{r oo-sample}
ca_tz <- "Etc/GMT+7"
bcg_begin <- lubridate::ymd_hms("2021-08-16 10:09:24.5", tz = ca_tz)
bcg_end <- lubridate::ymd_hms("2021-08-16 10:09:39", tz = ca_tz)
bcg_sample <- filter(oo_400hz, between(dt, bcg_begin, bcg_end))

@@

```{r oo-bcg-ecg, fig.cap="A: BCG, B: ECG, C: BPM comparison"}
# A: Surge + LSR
bcg_surge_lsr <- bcg_sample %>% 
  rename(Surge = surge_filt, LSR = surge_range) %>% 
  pivot_longer(c(Surge, LSR), 
               names_to = "var",
               values_to = "val")

@@

# C: BCG + ECG
pC <- ggplot(filter(bcg_sample, bcg_beat), aes(dt, bcg_bpm)) +
  geom_line(size = 0.2) +
  geom_line(aes(y = ecg_bpm), filter(bcg_sample, ecg_beat), size = 0.2) +
  geom_point(color = "blue") +
  geom_point(aes(y = ecg_bpm), filter(bcg_sample, ecg_beat), color = "red") +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = "Heart rate (bpm)") +
  expand_limits(x = range(bcg_sample$dt) + c(0, 1), y = c(30, 70)) +
  t

@@

```{r ecg_bcg_comparison}
ecg_bpm <- filter(bcg_sample, ecg_beat)$ecg_bpm %>% tail(-1)
bcg_bpm <- filter(bcg_sample, bcg_beat)$bcg_bpm %>% head(-1)
bpm_diff <- abs((bcg_bpm - ecg_bpm) / ecg_bpm * 100)
sprintf("%0.1f%% \u00B1 %0.1f%% (mean \u00B1 sd))", mean(bpm_diff), sd(bpm_diff))
```

```{r oo-bcg-acc, fig.cap="A: The BCG signal is clearest in surge (longitudinal axis acceleration). Solid line: surge bandpass filtered to [1.0 25.0] Hz. Dashed line: local range of surge in 0.25 s sliding window (smoothed). Blue dots indicate peak of the BCG I-wave. B: BCG signal is not as strong in heave (dorso-ventral axis) or sway (lateral axis) as in surge. The sway signal is likely due to movement of the chest cavity (see video S1). C: Some, but not all, heart beats are visible as peaks in the norm of the jerk vector."}
# A: Surge
pA <- ggplot(bcg_sample, aes(dt, surge_filt)) +
  geom_line(size = 0.2) +
  geom_point(data = filter(bcg_sample, bcg_beat), color = "blue") +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = bquote(Surge~(m~s^{-2}))) +
  expand_limits(x = max(bcg_sample$dt) + 1) +
  t

# B: Sway, heave
bcg_sway_heave <- bcg_sample %>% 
  rename(Sway = sway_filt, Heave = heave_filt) %>% 
  pivot_longer(c(Sway, Heave), 
               names_to = "axis",
               values_to = "acceleration")
pB <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
  geom_point(data = filter(bcg_sway_heave, bcg_beat), color = "blue") +
  geom_line(size = 0.2) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  facet_grid(axis ~ .) +
  labs(y = bquote(Acc~(m~s^{-2}))) +
  t

# C: Jerk
pC <- ggplot(bcg_sample, aes(dt, njerk_smooth)) +
  geom_line(size = 0.2) +
  geom_point(data = filter(bcg_sample, bcg_beat), color = "blue") +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = bquote(Jerk~(m~s^{-3}))) +
  t

cowplot::plot_grid(pA, pB, pC,
                   align = "v", 
                   axis = "lr",
                   nrow = 3, 
                   labels = "AUTO")
```

```{r bw-beats}
window = 1.8 # s
bw_elg <- bw180905_53_elg

# Apply filters as with killer whale data, but narrower band
bw_400hz <- bw180905_53_400hz %>% 
  mutate(
    # Pass-band filter acceleration
    across(surge:heave, 
           bandpass_filter, pass = c(1.0, 10.0), 
           .names = "{.col}_filt"),
    # Calculate jerk norm and apply smoother
    njerk = njerk(cbind(surge_filt, sway_filt, heave_filt), 
                  p = 4, n = window * fs),
    njerk_smooth = tma(njerk, window * fs),
    # Annotate regions
    rid_left = approx(bw_elg$start, bw_elg$region_id, dt, "constant")$y,
    rid_right = approx(bw_elg$stop, bw_elg$region_id, dt, "constant", yleft = 0)$y + 1,
    region_id = ifelse(rid_left == rid_right, rid_left, NA),
    # Find heart beats using jerk cue
    surge_range = local_range(surge_filt, window * fs),
    bcg_beat = find_beats_jerk(njerk_smooth, !is.na(region_id), window, fs)
  ) %>% 
  # Calculate heart rate within each region
  group_by(region_id) %>% 
  mutate(bcg_bpm = bpm(bcg_beat, dt)) %>% 
  ungroup() %>% 
  select(-c(rid_left, rid_right))
```

```{r bw-bcg}
bw_bcg_plot <- function(rid) {
  ca_tz <- "Etc/GMT+7"
  bcg_sample <- filter(bw_400hz, region_id == rid)

@@

  pA <- ggplot(bcg_surge_lsr, aes(dt, val, linetype = var)) +
    # geom_vline(aes(xintercept = dt),
    #            data = filter(bcg_surge_lsr, bcg_beat), 
    #            color = "blue", alpha = 0.5) +
    geom_line(size = 0.2) +
    geom_point(data = filter(bcg_surge_lsr, bcg_beat), color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    scale_linetype_manual(values = c(Surge = "solid", LSR = "dashed")) +
    labs(y = bquote(Surge~(m~s^{-2}))) +
    t +
    theme(legend.position = "none")

@@

  pB <- ggplot(filter(bcg_sample, bcg_beat), aes(dt, bcg_bpm)) +
    geom_line() +
    geom_point(color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    labs(y = "Heart rate (bpm)") +
    expand_limits(x = range(bcg_sample$dt), y = 0) +
    t

@@

  pC <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
    # geom_vline(aes(xintercept = dt),
    #            filter(bcg_sway_heave, bcg_beat), 
    #            color = "blue", alpha = 0.5) +
    geom_line(size = 0.2) +
    geom_point(data = filter(bcg_sway_heave, bcg_beat), color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    facet_grid(axis ~ .) +
    labs(y = bquote(Acc~(m~s^{-2}))) +
    t

@@

  pD <- ggplot(bcg_sample, aes(dt, njerk_smooth)) +
    geom_line(size = 0.2) +
    geom_point(data = filter(bcg_sample, bcg_beat), color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    labs(y = bquote(Jerk~(m~s^{-3}))) +
    expand_limits(y = 0) +
    t

  p <- cowplot::plot_grid(pA, pB, pC, pD,
                          align = "v", 
                          axis = "lr",
                          nrow = 4, 
                          rel_heights = c(1.5, 0.75, 1, 1),
                          labels = "AUTO")

@@

Figure \@ref(fig:oo-bcg-ecg) shows how we can have a caption and cross-reference for a plot.
